---
title: "00_all"
format: html
editor: visual
---

## Libraries

```{r}
library(GEOquery)
library(limma)
library(tidyverse)
library(edgeR)
library(TCGAbiolinks)
library(tidyverse)
library(stringr)
library(SummarizedExperiment)
library(patchwork)
library(reshape2)
```

# Data retrieval:

See the README.md file.

## GSE

```{r}
data <- getGEO(GEO = "GSE39833")[[1]]

exp_data <- exprs(data) #extracting expression values (Limma)

gpl <- Table(getGEO("GPL14767"))
```

### Tidying data

```{r}
data_frame <- as.data.frame(exp_data) |> 
  rownames_to_column(var = "ID")

gpl <- gpl |> 
  mutate(ID = as.character(ID))

mirna_exp <- data_frame |> 
  left_join(gpl, by = "ID") |> 
  select( ,-1) |> 
  relocate(where(is.numeric), .after = where(is.character))

metadata <- data@phenoData@data

```

### Saving datasets

```{r}
path <- "C:/Users/Laila/Desktop/R for Bio Data Science/group_04_project"

saveRDS(data_frame, file = paste0(path, "/data/data_load.RDs"))

exp_data <- as.data.frame(exp_data)
saveRDS(exp_data, file = paste0(path, "/data/gse_expressed_data.RDs"))

saveRDS(gpl, file = paste0(path, "/data/gse_chosen_meta_data.RDs"))

saveRDS(mirna_exp, file = paste0(path, "/data/gse_all_data.RDs"))

saveRDS(metadata, file = paste0(path, "/data/gse_all_meta_data.RDs"))


```

## TCGA

See the README.md file.

Obtaining the IDs of the patients for whom the RNA expression was registered:

```{r}
analyte <- read.delim("analyte.tsv")

RNA_samples <- subset(analyte, analyte_type == "RNA")

RNA_samples <- RNA_samples |>
  mutate(tissue_type = str_sub(sample_submitter_id, -3, -3)) |>
  mutate(
    type = case_when(tissue_type == 0 ~ "cancer",
                     (tissue_type == 1 | tissue_type == 2) ~ "normal",
                     .default = "other")
  ) |>
  arrange(type)
```

### Obtaining the IDs of interest

```{r}
id_cancer_patients_normal <- get_sample_ids(RNA_samples, sample_type = "normal", num_samples = 50)
id_cancer_patients_cancer <- get_sample_ids(RNA_samples, sample_type = "cancer", num_samples = 50)
id_cancer_patients <- c(id_cancer_patients_normal, id_cancer_patients_cancer)
```

## Retrieving datasets from TCGA

### miRNA

```{r}
miRNA_data <- retrieve_and_prepare_data(
  project = "TCGA-COAD",
  data_category = "Transcriptome Profiling",
  data_type = "miRNA Expression Quantification",
  workflow_type = "BCGSC miRNA Profiling",
  id_cancer_patients = id_cancer_patients,
  directory_prefix = "samples_miRNA"
)
```

## Fetch miRNA metadataset

```{r}
miRNA_metadata <- read.delim("clinical.tsv")
```

### mRNA

```{r}
mRNA_all <- retrieve_and_prepare_data(
  project = "TCGA-COAD",
  data_category = "Transcriptome Profiling",
  data_type = "Gene Expression Quantification",
  workflow_type = "STAR - Counts",
  id_cancer_patients = id_cancer_patients,
  directory_prefix = "samples_mRNA"
)
```

## References

```{r}
print(citation("TCGAbiolinks"), bibtex=TRUE)
```

## Save data files

```{r}
saveRDS(miRNA_data, file = paste0(own_path, "/group_04_project/data/miRNA_data_load.RDS"))
saveRDS(miRNA_metadata, file = paste0(own_path, "/group_04_project/data/miRNA_metadata_load.RDS"))
saveRDS(mRNA_all, file = paste0(own_path, "/group_04_project/data/mRNA_all_load.RDS"))
```

## Fetch dataset

```{r}
mRNA_all <- readRDS(paste0(own_path, "/group_04_project/data/mRNA_all_load.RDS"))
miRNA_data <- readRDS(paste0(own_path, "/group_04_project/data/miRNA_data_load.RDS"))
miRNA_metadata <- readRDS(paste0(own_path, "/group_04_project/data/miRNA_metadata_load.RDS"))
```

## Tidy miRNA data

```{r}
miRNA_data <- miRNA_data |>
  select(-starts_with("cross-mapped")) |>
  select(-starts_with("reads_per_million")) |>
  rename_with(~gsub("^read_count_", "", .x), starts_with("read_count_")) |>
  rename_all(~ substr(., 1, 16))
```

## Tidy miRNA metadata

```{r}
miRNA_metadata <- miRNA_metadata |>
  filter(case_submitter_id %in% substr(names(miRNA_data)[-1], 1, 12)) |>
  mutate(across(everything(), ~ifelse(. == "'--", NA, .))) |>
  select(-where(~all(is.na(.))),-treatment_type, -treatment_or_therapy)

miRNA_metadata <- unique(miRNA_metadata)

miRNA_metadata <- miRNA_metadata |>
  mutate(case_submitter_id = ifelse(
    case_submitter_id %in% substr(names(miRNA_data)[-1], 1, 12),
    names(miRNA_data)[-1],
    case_submitter_id))
```

## Alter column data types

```{r}
miRNA_metadata <- miRNA_metadata |>
  mutate(
    age_at_index = as.numeric(age_at_index),
    age_at_diagnosis = as.numeric(age_at_diagnosis),
    days_to_birth = as.numeric(days_to_birth),
    days_to_death = as.numeric(days_to_death),
    days_to_diagnosis = as.numeric(days_to_diagnosis),
    days_to_last_follow_up = as.numeric(days_to_last_follow_up),
    year_of_birth = as.numeric(year_of_birth),
    year_of_death = as.numeric(year_of_death),
    year_of_diagnosis = as.numeric(year_of_diagnosis)
         )
```

## Tidy mRNA data

```{r}
mRNA_data <- as.data.frame(assay(mRNA_all))
```

## Tidy mRNA metadataset

```{r}
mRNA_metadata <- as.data.frame(colData(mRNA_all))
```

## Save clean data files

```{r}
saveRDS(miRNA_data, file = paste0(own_path, "/group_04_project/data/miRNA_data_clean.RDS"))
saveRDS(miRNA_metadata, file = paste0(own_path, "/group_04_project/data/miRNA_metadata_clean.RDS"))
saveRDS(mRNA_data, file = paste0(own_path, "/group_04_project/data/mRNA_data_clean.RDS"))
saveRDS(mRNA_metadata, file = paste0(own_path, "/group_04_project/data/mRNA_metadata_clean.RDS"))
```

## Fetch datasets

```{r}
miRNA_metadata <- readRDS(paste0(own_path, "/group_04_project/data/miRNA_metadata_aug.RDS"))
mRNA_metadata <- readRDS(paste0(own_path, "/group_04_project/data/mRNA_metadata_aug.RDS"))
```

## Experiment description

#### Number og Cancer vs. Normal Samples

```{r}
my_colors <- c("Normal" = "cyan3", "Tumor" = "cyan4", "normal" = "cyan3", "cancer" = "cyan4")

plot1 <- ggplot(miRNA_metadata, aes(x = Status, fill = Status)) +
  geom_bar() +
  labs(x = "", y = "Frequency", title = "miRNA") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  guides(fill = FALSE) +
  scale_fill_manual(values = my_colors) +
  coord_cartesian(ylim = c(0, 75)) +
  facet_wrap(~ "Diagnosis", ncol = 2, scales = "free_x")

plot2 <- ggplot(mRNA_metadata, aes(x = Status, fill = Status)) +
  geom_bar() +
  labs(x = "", y = "", title = "mRNA") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  scale_fill_manual(values = my_colors) +
  coord_cartesian(ylim = c(0, 75)) +
  facet_wrap(~ "Diagnosis", ncol = 2, scales = "free_x")

cancer_vs_normal <- plot1 + plot2 + plot_layout(ncol = 2) +
  plot_annotation(title = "Frequency of Cases by Category", theme = theme(plot.title = element_text(hjust = 0.5)))

cancer_vs_normal

```

#### Gender distribution

```{r}
# Plot for miRNA_metadata with adjusted fill colors and y-axis limit
plot1 <- ggplot(miRNA_metadata, aes(x = gender, y = age_at_index, fill = gender)) + geom_violin() +
  labs(x = "", y = "Age", title = "miRNA") +
  guides(fill = FALSE) +
  facet_wrap(~ "Gender", ncol = 2, scales = "free_x") +
  geom_boxplot(width=0.1)

# Plot for mRNA_metadata with adjusted fill colors and y-axis limit
plot2 <- ggplot(mRNA_metadata, aes(x = gender, y = age_at_index, fill = gender)) + geom_violin() +
  labs(x = "", y = "", title = "mRNA") +
  facet_wrap(~ "Gender", ncol = 2, scales = "free_x") +
  geom_boxplot(width=0.1)

# Combine plots using patchwork and add a main title
gender_distribution <- plot1 + plot2 + plot_layout(ncol = 2) +
  plot_annotation(title = "Distribution of Age by Gender", theme = theme(plot.title = element_text(hjust = 0.5)))

gender_distribution
```

#### Cancer stages

```{r}
cancer_stage <- vctrs::vec_c(miRNA_metadata$ajcc_pathologic_stage, mRNA_metadata$ajcc_pathologic_stage)
cancer_stage_df <- data.frame(stage = unlist(cancer_stage))

cancer_stages <- ggplot(cancer_stage_df, aes(x = stage, fill = stage)) +
  geom_bar() +
  labs(x = "AJCC Pathologic Stage", y = "Frequency", title = "Frequency of AJCC Pathologic Stages by Status") +
  theme(axis.text.x = element_blank())

cancer_stages
```

## Save plots

```{r}
ggsave(file.path("../results/images/", "cancer_vs_normal.png"), cancer_vs_normal, width = 10, height = 6, units = "in")
ggsave(file.path("../results/images/", "gender_distribution.png"), gender_distribution, width = 10, height = 6, units = "in")
ggsave(file.path("../results/images/", "cancer_stages.png"), cancer_stages, width = 10, height = 6, units = "in")
```

# Data Analysis

## GSE

## Loading files

```{r}

exp_gse_results <- readRDS("data/gse_expressed_data.RDs")
mirna_exp <- readRDS("data/gse_all_data.RDs")


```

At first only necessary rows with gene_id and gse_results were selected. Then log2 transformation was done, with omitting negative values and 0's. Also box-plot is generated to see, whether data looks normalized or not

```{r}
exp_gse_results <- replace(exp_gse_results, exp_gse_results <= 0, NaN)

exp_log <- log2(exp_gse_results) # log2 transform

exp_log <- normalizeBetweenArrays(exp_log) # normalize gse_results

exp_log  <- exp_log [complete.cases(exp_log), ]

# boxplot generated as check if data is normalized, could not be done with ggplot
boxplot(exp_log, outline = FALSE,
        main = "Boxplot of Exosomal miRNA in Colorectal Samples",
        xlab = "", ylab = "Log2")

```

Next a matrix for our experiment was created, with first 11 samples being "Normal" and rest as "Tumor" group

```{r}

design <- data.frame(
  Normal = c(rep(1, 11), rep(0, 88)),
  Tumor = c(rep(0, 11), rep(1, 88))
)

design <- as.matrix(design)
```

Adfterwards we will cut off genes that are not expressed in more 1/3 of our samples

```{r}
cutoff <- median(exp_log)

is_expressed <- exp_log > cutoff

keep <- rowSums(is_expressed) > 33

diff_expressed_genes <- exp_log[keep, ]

```

Then we have to fit linear model and to make 2 subgroups - Tumor and Normal, and perform *empirical Bayes'* step to get statistics and p-values.\

```{r}
fit <- lmFit(diff_expressed_genes, design)

contrasts <- makeContrasts(Tumor - Normal, levels=design)

fit2 <- contrasts.fit(fit, contrasts)

fit2 <- eBayes(fit2)

topTable(fit2)
```

Now we will check which test we have to perform for our datset\

```{r}
decideTests(fit2)

table(decideTests(fit2))
```

We create array weights to fit our data into model, and once again apply *empirical Bayes' step*

```{r}
aw <- arrayWeights(diff_expressed_genes, design)

fit <- lmFit(diff_expressed_genes, design, cweights = aw)

contrasts <- makeContrasts(Tumor - Normal, levels=design)

fit2 <- contrasts.fit(fit, contrasts)

fit2 <- eBayes(fit2)

full_results <- topTable(fit2, number=Inf)

```

Results of our analysis are added to the data set and volcano plot was generated to show Diferentially Expressed genes

```{r}
full_results <- full_results |>
  tibble::rownames_to_column(var = "ID")

gse_results <- mirna_exp |>
  tibble::rownames_to_column(var = "ID") |>
  full_join(full_results, by = "ID")


gse_results$diffexpressed <- "NO"

gse_results$diffexpressed[gse_results$logFC > 1 & gse_results$adj.P.Val < 0.05] <- "UP"
gse_results$diffexpressed[gse_results$logFC < -1 & gse_results$adj.P.Val < 0.05] <- "DOWN"

gse_results$label <- NA
gse_results$label[gse_results$diffexpressed != "NO"] <- gse_results$GENE_NAME[gse_results$diffexpressed != "NO"] 

# Generate the plot with labels that avoid overlapping
gse_vol <- ggplot(data = gse_results, aes(x = logFC,
                               y = -log10(P.Value), 
                               col = diffexpressed, 
                               label = label)) + 
  theme_minimal() +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("UP" = "red", 
                                "DOWN" = "blue", 
                                "NO" = "gray"),
                     name = "Type of genes regulation") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  geom_text_repel(data = subset(gse_results, 
                                diffexpressed != "NO"), 
                  aes(label = label), 
                  box.padding = unit(0.2, "lines")) +
  labs(title = "Differential Gene Expression in Colorectal Cancer", 
       x = "log2 Fold Change", 
       y = "-log10(adj.P.Value)", 
       col = "Differential Expression")



print(gse_vol)
```

As the last step a file of only highly expressed genes is generated for merging purposes, and to create top 20 over expressed genes as an expression file

```{r}
high_expressed_genes <- gse_results |>
  filter(abs(logFC) > 1) |>
  filter(adj.P.Val < 0.05)

top_20_genes <- high_expressed_genes |>
  arrange(desc(abs(logFC))) |>
  slice_head(n = 20) |>
  mutate(miRNA_ID = ifelse(duplicated(miRNA_ID), 
                           paste0(miRNA_ID, "*"), 
                           miRNA_ID)) |>
  column_to_rownames(var = "miRNA_ID") |>
  select(starts_with("GS"))


```

## Saving Files

```{r}

saveRDS(gse_results, file = file.path(path, "gse_results.rds"))

saveRDS(high_expressed_genes, file = file.path(path, "gse_expressed_genes.rds"))

saveRDS(top_20_genes, file = file.path(path, "gse_top_20.rds"))

ggsave(file.path(path, "gse_vol_plot.pdf"), plot = gse_vol, width = 14, height = 8.4, units = "in")

```

## miRNA

## Set Working directory

```{r setup}
own_path <- "/Users/mathildebechgaard"
knitr::opts_knit$set(root.dir = paste0(own_path, "/group_04_project/data/"))
source(paste0(own_path, "/group_04_project/project_functions.R"))
```

## Fetch dataset

```{r}
miRNA_log_long <- readRDS(paste0(own_path, "/group_04_project/data/miRNA_aug.RDS"))
top_genes_stat <- readRDS(paste0(own_path, "/group_04_project/data/miRNA_top_genes_stat.RDS"))
miRNA_meta <- readRDS(paste0(own_path, "/group_04_project/data/mRNA_metadata_clean.RDS"))
```

Subsetting top 15 "log Fold Changed" genes for a better visualisation:

```{r}
top_genes_stat_subset <- top_genes_stat |>
  arrange(desc(abs(logFC))) |>
  head(n = 15)

top_genes <- top_genes_stat_subset$miRNA_ID
```

Uniting Stages in one group to facet the plots based on them later on.

```{r}
miRNA_meta <- miRNA_meta |>
 mutate(Stage = case_when(
    ajcc_pathologic_stage == "Stage I" | ajcc_pathologic_stage == "Stage IA" ~ "Stage I",
    ajcc_pathologic_stage == "Stage II" | ajcc_pathologic_stage == "Stage IIA" |
    ajcc_pathologic_stage == "Stage IIB" ~ "Stage II",
    ajcc_pathologic_stage == "Stage III" | ajcc_pathologic_stage == "Stage IIIA" |
    ajcc_pathologic_stage == "Stage IIIB" | ajcc_pathologic_stage == "Stage IIIC" ~ "Stage III",
    ajcc_pathologic_stage == "Stage IV" | ajcc_pathologic_stage == "Stage IVB" ~ "Stage IV"
  )) |>
  select(-ajcc_pathologic_stage) |> 
  tibble() |>
  select(-sample_submitter_id) |>
  dplyr::rename(TCGA_ID = "sample")
```

## Comparison of the expression between the stages

```{r}
miRNA_box <- miRNA_meta |> 
  select(TCGA_ID, Stage) |> 
  right_join(miRNA_log_long, by = 'TCGA_ID') |>
  set_tissue_status() |>
  ggplot(aes(x = Stage,
         y = log_reads,
         col = Status)) +
  geom_boxplot() +
  scale_color_manual(values = c("cancer" = "#f47c6c", 
                                "normal" = "#13b3ba")) + 
  labs(title = "Boxplot comparison of total miRNA expression between four cancer stages", color = "Tissue Type",
       x = "Cancer stage",
       y = "Log2 expression") +
  theme(plot.title = element_text(size=20),
        axis.title = element_text(size=15), 
        legend.text = element_text(size=15),
        legend.title = element_text(size=20))
```

## General comparison of the expression between cancer and normal tissues

-   Ideally, the samples should be taken from cancer and normal tissues of the same patients to get a proper overview of the expression difference. In our case, not all the samples had this comparison. Therefore, to make the conclusions based on the plots more precise, one should properly collect the expression data.

-   From the graph below, it can be seen that, in general, miRNAs expression is higher in cancer tissues.

```{r}
miRNA_expression <- miRNA_log_long |>
  filter(miRNA_ID %in% top_genes) |>
  set_tissue_status() |>
  ggplot(aes(x = miRNA_ID,
             y = log_reads,
             col = Status)) +
  geom_point() +
  labs(title = "Comparison of expression of 15 top differentially expressed miRNAs \nbetween cancer and normal TCGA samples", color = "Tissue Type",
       x = "miRNA IDs",
       y = "Log2 expression") +
  scale_color_manual(values = c("cancer" = "#f47c6c", 
                                "normal" = "#13b3ba")) +
  theme(axis.text.x = element_text(
    angle = 45,
    hjust = 1),
    plot.title = element_text(
      size = 12, 
      angle = 0)) +
  theme(plot.title = element_text(size=20),
        axis.title = element_text(size=15), 
        legend.text = element_text(size=15),
        legend.title = element_text(size=20))
```

## Volcano plot

-   In the following volcano plot, statistically significant values are highlighted in blue (the p-value is flipped, so it becomes more significant with the higher values). Positive LFC indicates higher expression in the cancer group compared to the normal tissues, that is in accordance with the plots above.

```{r}
# stating the significance based on the adjusted pvalues (FDR corrects for multiple comparisons)
top_genes_stat <- top_genes_stat |>
    mutate(
      significance = case_when(
        FDR > 0.05 ~ "no",
        FDR <= 0.05 ~ "yes"
      )
    )
```

```{r}
top_genes_stat <- top_genes_stat |>
  mutate(
    regulation = case_when(
      logFC > 1 ~ "Upregulated",
      logFC < -1 ~ "Downregulated",
      TRUE ~ "Non-significant"
    )
  )
```

```{r}
mirna_vol <- top_genes_stat |> 
  ggplot(aes(x = logFC,
             y = -log10(FDR),
             col = regulation)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("Upregulated" = "red", 
                                "Downregulated" = "blue", 
                                "Non-significant" = "gray"),
                     name = "Type of genes regulation") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  labs(x = "Log Fold Change",
       y = "-log10(FDR)",
       title = "Top Differentially Expressed miRNAs in TCGA-COAD Samples") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text_repel(data = top_genes_stat_subset,
                  aes(label = miRNA_ID),
                      color = "black") +
  theme(plot.title = element_text(size=20),
        axis.title = element_text(size=15), 
        legend.text = element_text(size=15),
        legend.title = element_text(size=20))
```

## Save plots

```{r}
ggsave(file.path(file = paste0(own_path, "/group_04_project/results/images/mirna_boxplot.png")), plot = miRNA_box, width = 14, height = 8, unit = "in")

ggsave(file.path(file = paste0(own_path, "/group_04_project/results/images/mirna_expression.png")), plot = miRNA_expression, width = 14, height = 8, unit = "in")

ggsave(file.path(file = paste0(own_path, "/group_04_project/results/images/mirna_vol.png")), plot = mirna_vol, width = 14, height = 8, unit = "in")
```

---
title: "mRNA_analysis"
format: html
editor: visual
---

## Set Working directory

```{r setup}
own_path <- "/Users/mathildebechgaard"
knitr::opts_knit$set(root.dir = paste0(own_path, "/group_04_project/data/"))
source(paste0(own_path, "/group_04_project/project_functions.R"))
```

## Fetch dataset

```{r}
mRNA_log_long <- readRDS(paste0(own_path, "/group_04_project/data/mRNA_aug.RDS"))
top_genes_stat <- readRDS(paste0(own_path, "/group_04_project/data/mRNA_top_genes_stat.RDS"))
mRNA_metadata <- readRDS(paste0(own_path, "/group_04_project/data/mRNA_metadata_clean.RDS"))
```

Subseting top 25 "log Fold Changed" genes for a better visualisation:

```{r}
top_genes_stat_subset <- top_genes_stat |>
  arrange(desc(abs(logFC))) |>
  head(n = 25)

top_genes <- top_genes_stat_subset$genes
```

## Stages comparison

Uniting Stages in one group to facet the plots based on them later on.

```{r}
mRNA_metadata <- mRNA_metadata |> 
  tibble() |>
  dplyr::rename(TCGA_ID = "barcode")
```

-   It can be concluded from the graph that the expression is higher in the normal tissues.

```{r}
mRNA_box <- mRNA_metadata |> 
  set_tissue_status() |>
   right_join(mRNA_log_long, by = 'TCGA_ID') |>
  ggplot(aes(x = gender,
         y = log_reads,
         col = Status)) +
  geom_boxplot() +
  scale_color_manual(values = c("cancer" = "#f47c6c", 
                                "normal" = "#13b3ba")) + 
  labs(title = "Boxplot comparison of total mRNA expression \nbetween genders", color = "Tissue Type",
       x = "Gender",
       y = "Log2 expression") +
  theme(plot.title = element_text(size=20),
        axis.title = element_text(size=15), 
        legend.text = element_text(size=15),
        legend.title = element_text(size=20))
```

## General expression comparison between cancer and normal tissues

-   From the graph below, it can be seen that, in general, mRNAs expression is higher in normal tissues for the genes where there is enough datapoints. However, there is not enough data for the normal tissues, that is seen from a little number of plotted points.

```{r}
mRNA_expression <- mRNA_log_long |>
  set_tissue_status() |>
  filter(genes %in% top_genes) |>
  ggplot(aes(x = genes,
             y = log_reads,
             col = Status)) +
  geom_point() +
  labs(title = "Comparison of expression of 25 top differentially expressed mRNAs between cancer and normal TCGA samples", color = "Tissue Type",
       x = "mRNA IDs",
       y = "Log2 expression") +
  theme(axis.text.x = element_text(
    angle = 45,
    hjust = 1),
    plot.title = element_text(
      size = 12, 
      angle = 0)) +
  theme(plot.title = element_text(size=20),
        axis.title = element_text(size=15), 
        legend.text = element_text(size=15),
        legend.title = element_text(size=20))
```

## Volcano plot

-   In the following volcano plot, statistically significant values are highlighted in blue (the p-value is flipped, so it becomes more significant with the higher values). Positive LFC indicates higher expression in the cancer group compared to the normal tissues, that is in accordance with the plot above.

```{r}
# stating the significance based on the adjusted pvalues (FDR corrects for multiple comparisons)
top_genes_stat <- top_genes_stat |>
    mutate(
      significance = case_when(
        FDR > 0.05 ~ "no",
        FDR <= 0.05 ~ "yes"
      )
    )
```

```{r}
# defining the over-/underexpression character of genes
top_genes_stat <- top_genes_stat |>
  mutate(
    regulation = case_when(
      logFC > 1 ~ "Upregulated",
      logFC < -1 ~ "Downregulated",
      TRUE ~ "Non-significant"
    )
  )
```

```{r}
mrna_vol <- top_genes_stat |> 
  ggplot(aes(x = logFC,
             y = -log10(FDR),
             col = regulation)) +
  geom_point() +
  scale_color_manual(values = c("Upregulated" = "red", 
                                "Downregulated" = "blue", 
                                "Non-significant" = "gray"),
                     name = "Type of genes regulation") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  labs(x = "Log Fold Change",
       y = "-log10(FDR)",
       title = "Top Differentially Expressed mRNAs in TCGA-COAD Samples") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_text_repel(data = head(top_genes_stat_subset, 15),
    aes(x = logFC, y = -log10(FDR), label = genes),
    color = "black") +
  theme(plot.title = element_text(size=20),
        axis.title = element_text(size=15), 
        legend.text = element_text(size=15),
        legend.title = element_text(size=20))
```

## Save plots

```{r}
ggsave(file.path(file = paste0(own_path, "/group_04_project/results/images/mrna_boxplot.png")), plot = mRNA_box, width = 14, height = 8, unit = "in")

ggsave(file.path(file = paste0(own_path, "/group_04_project/results/images/mrna_expression.png")), plot = mRNA_expression, width = 14, height = 8, unit = "in")

ggsave(file.path(file = paste0(own_path, "/group_04_project/results/images/mrna_vol.png")), plot = mrna_vol, width = 14, height = 8, unit = "in")
```

---
title: "TCGA_topgenes_heatmap"
format: html
editor: visual
---

## Set Working directory

```{r setup}
own_path <- "/Users/mathildebechgaard"
knitr::opts_knit$set(root.dir = paste0(own_path, "/group_04_project/data/"))
```

## Load datasets

```{r}
top_genes_stat_miRNA <- readRDS(paste0(own_path, "/group_04_project/data/miRNA_top_genes_stat.RDS"))
miRNA_data <- readRDS(paste0(own_path, "/group_04_project/data/miRNA_data_clean.RDS"))

top_genes_stat_mRNA <- readRDS(paste0(own_path, "/group_04_project/data/mRNA_top_genes_stat.RDS"))
mRNA_data <- readRDS(paste0(own_path, "/group_04_project/data/mRNA_data_clean.RDS"))
```

```{r}
mRNA_data <- mRNA_data |>
  rownames_to_column(var = "genes") |>
  tibble()
```

Function to choose top `n_genes` from the datasets.

```{r}
top_genes_choosing <- function(stat_data, 
                             id_column, 
                             n_genes,
                             expr_data) {
  top <- stat_data |>
  mutate(diff_col = case_when(
    logFC > 0.6 & FDR < 0.05 ~ "UP",
    logFC < -0.6 & FDR < 0.05 ~ "DOWN",
    TRUE ~ "NO"
  )) |> 
  arrange(desc(abs(logFC))) |> 
  select({{ id_column }}) |>
  head(n_genes)
  
  top <- top |> pull({{ id_column }})
  
  top_expression <- expr_data |>
  filter({{ id_column }} %in% top) |>
  head(n_genes)
  
  return(top_expression)
}
```

```{r}
miRNA_top_20_expression <- top_genes_choosing(top_genes_stat_miRNA, 
                   id_column = miRNA_ID, 
                   n_genes = 20, 
                   expr_data = miRNA_data)
```

```{r}
mRNA_top_20_expression <- top_genes_choosing(top_genes_stat_mRNA, 
                   id_column = genes, 
                   n_genes = 20, 
                   expr_data = mRNA_data)
```

```{r}
saveRDS(mRNA_top_20_expression, file = paste0(own_path, "/group_04_project/results/mRNA_top_expression.RDs"))
saveRDS(miRNA_top_20_expression, file = paste0(own_path, "/group_04_project/results/miRNA_top_expression.RDs"))
```

---
title: "TCGA_Heatmap"
format: html
editor: visual
---

## Set Working directory

```{r setup}
own_path <- "/Users/mathildebechgaard"
knitr::opts_knit$set(root.dir = paste0(own_path, "/group_04_project/data/"))
```

## Fetch dataset

```{r}
gse_miRNA <- readRDS(paste0(own_path, "/group_04_project/results/gse_top_20.rds"))
TCGA_miRNA <- readRDS(paste0(own_path, "/group_04_project/results/miRNA_top_expression.rds"))
TCGA_mRNA <- readRDS(paste0(own_path, "/group_04_project/results/mRNA_top_expression.rds"))
```

## Transform data for visualization

```{r}
GSE_miRNA <- gse_miRNA |>
  mutate(across(where(is.numeric), ~log2(. + 1)))

TCGA_miRNA <- TCGA_miRNA |>
  mutate(across(where(is.numeric), ~log2(. + 1)))
rownames(TCGA_miRNA) <- TCGA_miRNA[, 1]
TCGA_miRNA <- TCGA_miRNA[, -1]
TCGA_miRNA <- TCGA_miRNA[, colSums(is.na(TCGA_miRNA)) != nrow(TCGA_miRNA), drop = FALSE]

TCGA_mRNA <- as.data.frame(TCGA_mRNA) |>
  mutate(across(where(is.numeric), ~log2(. + 1)))
rownames(TCGA_mRNA) <- TCGA_mRNA[, 1]
TCGA_mRNA <- TCGA_mRNA[, -1]
```

## Transpose dataframes to fit geom_tile

```{r}
GSE_miRNA <- t(GSE_miRNA)
TCGA_miRNA <- t(TCGA_miRNA)
TCGA_mRNA <- t(TCGA_mRNA)
```

## miRNA Heatmap

```{r}
GSE_miRNA_heatmap <- ggplot(data = melt(GSE_miRNA), aes(x = Var2, y = Var1, fill = value)) +
  geom_tile() +
  labs(x = "Genes", y = "Patients", title = "Heatmap of Expression Values") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.text.y = element_blank())
GSE_miRNA_heatmap + scale_fill_gradient(limits = c(-10, 15))
```

## miRNA Heatmap

```{r}
TCGA_miRNA_heatmap <- ggplot(data = melt(TCGA_miRNA), aes(x = Var2, y = Var1, fill = value)) +
  geom_tile() +
  labs(x = "Genes", y = "Patients", title = "Heatmap of Expression Values") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.text.y = element_blank())
TCGA_miRNA_heatmap + scale_fill_gradient(limits = c(-10, 15))
```

## mRNA Heatmap

```{r}
TCGA_mRNA_heatmap <- ggplot(data = melt(TCGA_mRNA), aes(x = Var2, y = Var1, fill = value)) +
  geom_tile() +
  labs(x = "Genes", y = "Patients", title = "Heatmap of Expression Values") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.text.y = element_blank())
TCGA_mRNA_heatmap + scale_fill_gradient(limits = c(-10, 15))
```

## Save plots

```{r}
ggsave(file.path("../results/images/", "GSE_miRNA_heatmap.png"), GSE_miRNA_heatmap, width = 10, height = 6, units = "in")
ggsave(file.path("../results/images/", "TCGA_miRNA_heatmap.png"), TCGA_miRNA_heatmap, width = 10, height = 6, units = "in")
ggsave(file.path("../results/images/", "TCGA_mRNA_heatmap.png"), TCGA_mRNA_heatmap, width = 10, height = 6, units = "in")
```

## Clean the environment

```{r}
rm(list = ls())
```
