---
title: "GSE_data_analysis"
format: html
editor: visual
---

## Libraries

```{r}
library(tidyverse)
library(limma)
```

## Loading Data

```{r}

```

## Data analysis

at first only necessary rows with gene_id and data were selected. Then log2 transformation was done, with omiting negative values and 0's

```{r}
exp_data[which(exp_data <= 0)] <- NaN
exp_log <- log2(exp_data) # log2 transform

exp_log <- normalizeBetweenArrays(exp_log) # normalize data

exp_log  <- exp_log [complete.cases(exp_log), ]

boxplot(exp_log, outline = FALSE)

```

```{r}

design <- data.frame(
  Normal = c(rep(1, 11), rep(0, 88)),
  Tumour = c(rep(0, 11), rep(1, 88))
)

design <- as.matrix(design)
```

next we will cut off genes that are not expressed in most of the samples, in our case 97

```{r}
cutoff <- median(exp_log)

is_expressed <- exp_log > cutoff

keep <- rowSums(is_expressed) > 5

diff_expressed_genes <- exp_log[keep, ]

```

Next step is to fit linear model and to make 2 subgroups - Tumour and Normal, and perform *empirical Bayes'* step to get statistics and p-values.\
\

```{r}
fit <- lmFit(diff_expressed_genes, design)

contrasts <- makeContrasts(Tumour - Normal, levels=design)

fit2 <- contrasts.fit(fit, contrasts)

fit2 <- eBayes(fit2)

topTable(fit2)
```

Now we will check which test we have to perform for our datset\
\

```{r}
decideTests(fit2)
table(decideTests(fit2))
```

```{r}
aw <- arrayWeights(diff_expressed_genes, design)
fit <- lmFit(diff_expressed_genes, design,cweights = aw)
contrasts <- makeContrasts(Tumour - Normal, levels=design)
fit2 <- contrasts.fit(fit, contrasts)
fit2 <- eBayes(fit2)
full_results <- topTable(fit2, number=Inf)

```

```{r}
full_results <- full_results |>
  tibble::rownames_to_column(var = "ID")

mirna_results <- mirna_exp |>
  tibble::rownames_to_column(var = "ID") |>
  inner_join(full_results, by = "ID")

p_cutoff <- 0.05
fc_cutoff <- 1



full_results |>
  mutate(Significant = adj.P.Val < p_cutoff & abs(logFC) > fc_cutoff) |>
  ggplot(aes(x = logFC, y = -log10(adj.P.Val))) +
  geom_point(data = subset(full_results, logFC < 0), col = "blue", alpha = 0.5) +
  geom_point(data = subset(full_results, logFC > 0), col = "red", alpha = 0.5) +
  geom_point(data = subset(full_results, logFC >= -1 & logFC <= 1), col = "gray") +
  geom_vline(xintercept = c(-fc_cutoff, fc_cutoff), linetype = "dashed") +
  geom_hline(yintercept = -log10(p_cutoff), linetype = "dashed") +
  geom_text(
    data = mirna_results,
    aes(label = ifelse(abs(logFC) > 1.5, as.character(GENE_NAME), '')),
    vjust = 0, hjust = 0, check_overlap = FALSE, size = 1.5)+
geom_text_repel(data = subset(data, diffcol != "NO"), aes(label = label), 
               box.padding = unit(0.2, "lines")) +



```

```{r}
high_expressed_genes <- mirna_results |>
  filter(abs(logFC) > 1) |>
  slice(-14) |>
  inner_join(mirna_exp, by = "GENE_NAME", copy = FALSE) |>
  distinct(GENE_NAME, .keep_all = FALSE)

target_mirna <- mirna_exp |>
  filter(GENE_NAME %in% high_expressed_genes$GENE_NAME)|>
  distinct(GENE_NAME, .keep_all = TRUE)
```

## Deleting files

```{r}
objects_to_keep <- ls(pattern = "target_mirna")
rm(list = setdiff(ls(), objects_to_keep))
```
